---
title: "Histones and Apoptosis"
author: "Claudia Feng"
date:  "`r format(Sys.Date(),'%e %B, %Y')`"
output: BiocStyle::html_document
params:
  date: "2022-08-15"
  home_folder : "/lustre/scratch123/hgi/teams/parts/cf14/crispr_scrnaseq_hipsci/"
  section_name : "12c_highly_correlated_downstream_genes"
  complex_of_interest : "histones_and_apoptosis"
  experiment_name : "Magpie"
---

```{r, echo = F}
HomeFolder <- params$home_folder
ExperimentName <- params$experiment_name
date <- params$date
section_name <- params$section_name
network_of_interest <- params$network_of_interest
```


```{r prep, warning=FALSE, message=FALSE, echo = F}
# to render pngs in html
library(Cairo)
knitr::opts_chunk$set(fig.path = paste0(plotsdir, "/", tolower(gsub( x = params$complex_of_interest, pattern = " ", replacement = '_')), "/"), dev=c("CairoPNG", "pdf"))
print(params)
```
 
# Overlap Between Histones and the TP53 Pathway

## UMAP

```{r highlighted_umap, warning = F, echo = F, fig.width = 6, fig.height = 6 }
ggplot(umap_coords, aes(x = umap_1, y = umap_2, 
                        col = ifelse(umap_2 > 5, "y", "n"))) + 
  xlab("UMAP_1") + ylab("UMAP_2") + 
  scale_color_manual(values = c('y' = 'red', 'n' = 'gray')) + 
  geom_point(size = 0.4) + theme_bw()  + theme(legend.position = 'none')

```
## Zoomed-in UMAP

```{r highlighted_umap, warning = F, echo = F, fig.width = 6, fig.height = 6 }
ggplot(umap_coords, aes(x = umap_1, y = umap_2, 
                        label = gene)) + 
  xlab("UMAP_1") + ylab("UMAP_2") + 
  xlim(c(2, 2.3)) + ylim(c(5.5,7)) + 
  ggrepel::geom_text_repel(max.overlaps = Inf) + 
  geom_point() + theme_bw() 

```

Genes of interest are:

```{r, echo = F}
genes_of_interest <- umap_coords %>%
  dplyr::filter(umap_2 > 5.5) %>%
  .$gene %>% sort()
print("All genes of interest:")
cat(paste0(genes_of_interest, collapse = '\n'))


histone_genes <- umap_coords %>%
  dplyr::filter(umap_2 > 5) %>%
  .$gene %>% grep(pattern = 'H1', value = T)
print("Histone genes:")
cat(paste0(histone_genes, collapse = '\n'))
apoptosis_genes <- umap_coords %>%
  dplyr::filter(umap_2 < 6 & umap_2 > 5.5) %>%
  .$gene
print("Apoptosis genes:")
cat(paste0(apoptosis_genes, collapse = '\n'))

```

## Heatmap

### Comparison to iPSC co-expression

```{r heatmap_vs_ipsc_coexpression, fig.width = 5.25, fig.height=4.5, echo = F}

downstream_genes2plot <- genes_of_interest
sc_coexpression_genes <- sc_coexpression %>%
  dplyr::filter(gene_1 %in% downstream_genes2plot & gene_2 %in% downstream_genes2plot)
downstream_genes2plot <- sort(unique(c(sc_coexpression_genes$gene_1, sc_coexpression_genes$gene_2)))

df4heatmap_iPSCs <- downstream_downstream_cor_anno %>%
  dplyr::select(c('downstream_gene_1', 'downstream_gene_2', 'iPSC_cor')) %>%
  dplyr::filter(downstream_gene_1 %in% downstream_genes2plot & downstream_gene_2 %in% downstream_genes2plot) %>%
  reshape2::dcast(downstream_gene_1 ~ downstream_gene_2, value.var = 'iPSC_cor') %>%
  column_to_rownames('downstream_gene_1') %>%
  as.matrix()

df4heatmap_sc_coexpression <- matrix(1, nrow = length(downstream_genes2plot), ncol = length(downstream_genes2plot))
colnames(df4heatmap_sc_coexpression) <- row.names(df4heatmap_sc_coexpression) <- downstream_genes2plot
for (i in 1:length(downstream_genes2plot)){
  for (j in 1:i){
    if (i != j){
      df4heatmap_sc_coexpression[downstream_genes2plot[i],downstream_genes2plot[j]] <- df4heatmap_sc_coexpression[downstream_genes2plot[j],downstream_genes2plot[i]] <- sc_coexpression_genes %>% dplyr::filter(gene_1 == downstream_genes2plot[j] & gene_2 ==downstream_genes2plot[i]) %>% .$correlation
    }
    
  }
}

double_heatmap(cor_mat1 = df4heatmap_iPSCs, cor_mat2 = df4heatmap_sc_coexpression,
               min_c = -0.6, max_c = 0.6,
               border_col = NA,
           treeheight_row = 0, treeheight_col = 0)
```

### Comparison to K562 correlation

```{r heatmap_vs_k562_correlation, fig.width = 5.25, fig.height=4.5, echo = F}
downstream_genes2plot <- genes_of_interest
K562_genes <- downstream_downstream_cor_anno %>% 
  dplyr::filter(!(is.na(K562_gwps_cor))) %>%
  .$downstream_gene_1 %>%
  unique() %>% sort()
downstream_genes2plot <- intersect(downstream_genes2plot, K562_genes)

df4heatmap_iPSCs <- downstream_downstream_cor_anno %>%
  dplyr::select(c('downstream_gene_1', 'downstream_gene_2', 'iPSC_cor')) %>%
  dplyr::filter(downstream_gene_1 %in% downstream_genes2plot & downstream_gene_2 %in% downstream_genes2plot) %>%
  reshape2::dcast(downstream_gene_1 ~ downstream_gene_2, value.var = 'iPSC_cor') %>%
  column_to_rownames('downstream_gene_1') %>%
  as.matrix()
df4heatmap_K562 <- downstream_downstream_cor_anno %>%
  dplyr::select(c('downstream_gene_1', 'downstream_gene_2', 'K562_gwps_cor')) %>%
  dplyr::filter(downstream_gene_1 %in% downstream_genes2plot & downstream_gene_2 %in% downstream_genes2plot) %>%
  reshape2::dcast(downstream_gene_1 ~ downstream_gene_2, value.var = 'K562_gwps_cor') %>%
  column_to_rownames('downstream_gene_1') %>%
  as.matrix()

double_heatmap(cor_mat1 = df4heatmap_iPSCs, cor_mat2 = df4heatmap_K562,
               min_c = -0.6, max_c = 0.6,
               border_col = NA,
           treeheight_row = 0, treeheight_col = 0)
```

## Overlapping Targets
```{r}
targets_that_de <- magpie_lfcs_by_downstream[c(apoptosis_genes, histone_genes)] %>%
  bind_rows() %>% as.data.frame() %>%
  group_by(target) %>%
  summarize(
    n_deg = sum(pval_adj < sig_pval_thresh ),
    n_apoptosis_deg = sum(pval_adj < sig_pval_thresh & downstream_gene_name %in% apoptosis_genes),
    n_histone_deg = sum(pval_adj < sig_pval_thresh & downstream_gene_name %in% histone_genes),
    )
```

Slim pickings for common down-regulating targets.

## Low correlations

Check if the similarity between the two clusters is actually that convincing.

```{r tp53_hist_high_cor_scatters, echo = F, fig.width = 8.5, fig.height=8, warning = F, message = F, echo = F}

#dg1 <- "BBC3"; dg2 <- "HIST1H1E"
downstream_genes_in_umap <- downstream_genes2include
top_correlates <- downstream_downstream_cor_anno %>%
  dplyr::filter(downstream_gene_1 %in% apoptosis_genes & downstream_gene_2 %in% histone_genes) %>%
  slice_max(abs(iPSC_cor), n = 4)

plotlist <- mapply(1:dim(top_correlates)[1], FUN = function(i){
  dg1 <- top_correlates$downstream_gene_1[i]; dg2 <- top_correlates$downstream_gene_2[i]
  df4plot <- data.frame(
    dg1 = magpie_lfcs_by_downstream[[dg1]]$lfc,
    dg2 = magpie_lfcs_by_downstream[[dg2]]$lfc,
    target = magpie_lfcs_by_downstream[[dg2]]$target
  )
  p <- ggplot(df4plot, aes(x = dg1, y = dg2,
                    label = ifelse((abs(dg1) > sig_abs_lfc_thresh & abs(dg2) > sig_abs_lfc_thresh), target, ''))) +
    ggrepel::geom_text_repel(max.overlaps = 20) + 
    xlab(dg1) + ylab(dg2) + 
    geom_point(size =0.4) + theme_bw()
  return(p)
}, SIMPLIFY = F)
ggarrange(plotlist = plotlist)

```

I'm really not that convinced. Try subsetting to just the genes we considered in the heatmap?


```{r tp53_hist_high_cor_scatters_det_only, echo = F, fig.width = 8.5, fig.height=8, warning = F, message = F, echo = F}

#dg1 <- "BBC3"; dg2 <- "HIST1H1E"
target_genes_in_umap <- targets_that_de
top_correlates <- downstream_downstream_cor_anno %>%
  dplyr::filter(downstream_gene_1 %in% apoptosis_genes & downstream_gene_2 %in% histone_genes) %>%
  slice_max(abs(iPSC_cor), n = 4)

plotlist <- mapply(1:dim(top_correlates)[1], FUN = function(i){
  dg1 <- top_correlates$downstream_gene_1[i]; dg2 <- top_correlates$downstream_gene_2[i]
  df4plot <- data.frame(
    dg1 = magpie_lfcs_by_downstream[[dg1]]$lfc,
    dg2 = magpie_lfcs_by_downstream[[dg2]]$lfc,
    target = magpie_lfcs_by_downstream[[dg2]]$target
  ) %>% dplyr::filter(target %in% target_genes_in_umap)
  p <- ggplot(df4plot, aes(x = dg1, y = dg2,
                    label = ifelse((abs(dg1) > sig_abs_lfc_thresh & abs(dg2) > sig_abs_lfc_thresh), target, ''))) +
    ggrepel::geom_text_repel(max.overlaps = 20) + 
    xlab(dg1) + ylab(dg2) + 
    geom_point(size =0.4) + theme_bw()
  return(p)
}, SIMPLIFY = F)
ggarrange(plotlist = plotlist)

```

Lol no. 

Do they correlate well with CDK1?

```{r tp53_hist_low_cor_scatters, echo = F, fig.width = 8.5, fig.height=8, warning = F, message = F, echo = F}

#dg1 <- "BBC3"; dg2 <- "HIST1H1E"

top_correlates <- downstream_downstream_cor_anno %>%
  dplyr::filter(downstream_gene_1 %in% c(apoptosis_genes, histone_genes) & downstream_gene_2 %in% "CDK1") %>%
  slice_max(abs(iPSC_cor), n = 4)

plotlist <- mapply(1:dim(top_correlates)[1], FUN = function(i){
  dg1 <- top_correlates$downstream_gene_1[i]; dg2 <- top_correlates$downstream_gene_2[i]
  df4plot <- data.frame(
    dg1 = magpie_lfcs_by_downstream[[dg1]]$lfc,
    dg2 = magpie_lfcs_by_downstream[[dg2]]$lfc,
    target = magpie_lfcs_by_downstream[[dg2]]$target
  )
  p <- ggplot(df4plot, aes(x = dg1, y = dg2,
                    label = ifelse((abs(dg1) > sig_abs_lfc_thresh & abs(dg2) > sig_abs_lfc_thresh), target, ''))) +
    ggrepel::geom_text_repel(max.overlaps = 20) + 
    xlab(dg1) + ylab(dg2) + 
    geom_point(size =0.4) + theme_bw()
  return(p)
}, SIMPLIFY = F)
ggarrange(plotlist = plotlist)

```

Still not convincing. I wonder what's up with the correlation.


# Histones

## Choose Genes
### Low correlations

Check that the lowly correlated genes are actually convincing.

```{r correlation_between_clusters, echo = F,warning = F, message = F, fig.width = 10, fig.height = 9}

#dg1 <- "BBC3"; dg2 <- "HIST1H1E"
top_correlates <- downstream_downstream_cor_anno %>%
  dplyr::filter(downstream_gene_1 %in% genes_of_interest & downstream_gene_2 %in% genes_of_interest & downstream_gene_1 < downstream_gene_2) %>%
  slice_min(iPSC_cor, n = 4)

plotlist <- mapply(1:dim(top_correlates)[1], FUN = function(i){
  dg1 <- top_correlates$downstream_gene_1[i]; dg2 <- top_correlates$downstream_gene_2[i]
  df4plot <- data.frame(
    dg1 = magpie_lfcs_by_downstream[[dg1]]$lfc,
    dg2 = magpie_lfcs_by_downstream[[dg2]]$lfc,
    target = magpie_lfcs_by_downstream[[dg2]]$target
  )
  p <- ggplot(df4plot, aes(x = dg1, y = dg2,
                    label = ifelse((abs(dg1) > sig_abs_lfc_thresh & abs(dg2) > sig_abs_lfc_thresh), target, ''))) +
    ggrepel::geom_text_repel(max.overlaps = 20) + 
    xlab(dg1) + ylab(dg2) + 
    geom_point(size =0.4) + theme_bw()
  return(p)
}, SIMPLIFY = F)
ggarrange(plotlist = plotlist)

```


### Other Correlated Genes

```{r dist_of_correlation_across_clusters, echo = F}

df4plot <- downstream_downstream_cor_anno %>%
  dplyr::filter(downstream_gene_1 %in% genes_of_interest) %>%
  mutate(in_cluster = ifelse(downstream_gene_2 %in% genes_of_interest, 'in_cluster', 'outside_cluster'))

ggplot(df4plot %>%
         dplyr::filter(downstream_gene_1 != downstream_gene_2), aes(x = iPSC_cor, fill = in_cluster)) + 
  scale_fill_manual('', values = c("in_cluster" = downstream_col, 'outside_cluster' = 'gray')) + 
  xlab("Correlation") +ylab("") + 
  geom_density(alpha = 0.4) + theme_bw() + theme(legend.position = 'none')

#claudia <- df4plot %>% dplyr::filter(in_cluster == 'outside_cluster')
```




## Targets Driving Correlation

## Correlation Plots

There really aren't that many genes, so plot the correlation for all of them.

```{r hist_low_cor_scatters, echo = F, fig.width = 8.5, fig.height=8, warning = F, message = F, echo = F}

dg1 <- "HIST1H1E"; dg2 <- "HIST1H1B"

top_correlates <- downstream_downstream_cor_anno %>%
  dplyr::filter(downstream_gene_1 %in% genes_of_interest & downstream_gene_2 %in% genes_of_interest & downstream_gene_1 < downstream_gene_2) %>%
  slice_max(iPSC_cor, n = 4)

plotlist <- mapply(1:dim(top_correlates)[1], FUN = function(i){
  dg1 <- top_correlates$downstream_gene_1[i]; dg2 <- top_correlates$downstream_gene_2[i]
  df4plot <- data.frame(
    dg1 = magpie_lfcs_by_downstream[[dg1]]$lfc,
    dg2 = magpie_lfcs_by_downstream[[dg2]]$lfc,
    target = magpie_lfcs_by_downstream[[dg2]]$target
  )
  p <- ggplot(df4plot, aes(x = dg1, y = dg2,
                    label = ifelse((abs(dg1) > sig_abs_lfc_thresh & abs(dg2) > sig_abs_lfc_thresh), target, ''))) +
    ggrepel::geom_text_repel(max.overlaps = 20) + 
    xlab(dg1) + ylab(dg2) + 
    geom_point(size =0.4) + theme_bw()
  return(p)
}, SIMPLIFY = F)
ggarrange(plotlist = plotlist)

```

## Top Targets

Seems like CDH2 is universally down-expressed and CDH8 causes up-expression. Make it official with actual counts.

```{r, echo = F, warning = F, message = F, fig.width = 12, fig.height = 4}
target_de_counts <- magpie_lfcs %>%
  bind_rows() %>% as.data.frame() %>%
  #dplyr::filter(pval_adj < sig_pval_thresh) %>%
  group_by(target) %>%
  summarize(
    n_deg = sum(pval_adj < sig_pval_thresh),
    n_upregulated = length(which(lfc > 0 & pval_adj < sig_pval_thresh)),
    n_downregulated = length(which(lfc < 0 & pval_adj < sig_pval_thresh)),
    n_deg_in_cluster = sum(downstream_gene_name %in% histone_genes & pval_adj < sig_pval_thresh),
    n_downregulated_in_cluster =length(which(downstream_gene_name %in% histone_genes & lfc < 0 & pval_adj < sig_pval_thresh)),
    n_upregulated_in_cluster = length(which(downstream_gene_name %in% histone_genes & lfc > 0 & pval_adj < sig_pval_thresh))
  ) 
```

I didn't think this was all that great because there weren't a lot of genes that up/down-regulate the histones. My guess is that we're under-powered here. Maybe just do a heatmap

```{r}
top_genes <- magpie_lfcs %>%
  bind_rows() %>% as.data.frame() %>%
  dplyr::filter(downstream_gene_name %in% ) %>%
  group_by(target) %>%
  slice_max(abs(lfc))
```



# Save Files

```{r}
fwrite(target_de_counts, paste0(outdir, '/', tolower(gsub( x = params$complex_of_interest, pattern = " ", replacement = '_')), date, "_targets_that_cause_de.tsv"), sep = '\t')
```


# TP53

## Functional Enrichment of Genes

Check for functional enrichment of genes in the cluster

```{r}
gprofiler_res <- gprofiler2::gost(apoptosis_genes)$result
```

## Genes that Up/Downregulate the Pathway

```{r, echo = F}
targets_that_de <- magpie_lfcs_by_downstream[apoptosis_genes] %>%
  bind_rows() %>% as.data.frame() %>%
  dplyr::filter(target != downstream_gene_name) %>%
  group_by(target) %>%
  summarize(
    n_deg = sum(pval_adj < sig_pval_thresh ),
    n_upregulated = sum(pval_adj < sig_pval_thresh  & lfc > 0),
    n_downregulated = sum(pval_adj < sig_pval_thresh  & lfc < 0),
    n_downregulated_excl_target = sum(pval_adj < sig_pval_thresh  & lfc < 0 & target != downstream_gene_name),
    n_downstream_in_cluster = length(apoptosis_genes)
  ) %>%
  mutate(
    num_targets_total = target_meta$n_downstream_excl_target[match(target, target_meta$gene)],
    is_tf = target %in% tfs$source,
    is_tf_target = target %in% tfs$target,
    in_cluster = target %in% apoptosis_genes
  ) %>%
  dplyr::filter(!(target %in% c(off_targets$target_gene, off_targets$off_target_gene))) %>%
  dplyr::mutate(
    sensitivity = n_deg/num_targets_total,
    precision = n_deg/length(apoptosis_genes)
  ) %>% dplyr::filter(sensitivity > 0.2 & precision > 0.2 & !(target %in% genes2exclude))
targets_that_de <- left_join(targets_that_de, target_meta %>% dplyr::select(c("target" = "gene", contains("lfc"))))
to_save = list(
  targets_that_de = targets_that_de,
  genes_of_interest = apoptosis_genes
)
saveRDS(to_save, paste0(outdir, '/', date, '_targets_that_cause_apoptosis.RDS'))

ggplot(targets_that_de, aes(x = n_upregulated, y = pre_day10_lfc_mean)) + 
  geom_point()
```

Do the genes that 

## TFs with Targets in the Pathway

```{r}
tfs_with_targets <- tfs %>%
  group_by(source) %>%
  summarize(n_targets = n(),
            n_targets_in_pathway = sum(target %in% apoptosis_genes)) %>%
  dplyr::filter()
```


## Target-Downstream Effects

# Overlapping Effects

## 

# Session info

```{r session_info}
sessionInfo()
```