---
title: "The Cytosolic Ribosome"
author: "Claudia Feng"
date:  "`r format(Sys.Date(),'%e %B, %Y')`"
output: BiocStyle::html_document
params:
  date: "2022-08-15"
  home_folder : "/lustre/scratch123/hgi/teams/parts/cf14/crispr_scrnaseq_hipsci/"
  section_name : "12c_highly_correlated_downstream_genes"
  complex_of_interest : "apoptosis_tp53"
  experiment_name : "Magpie"
---

```{r, echo = F}
HomeFolder <- params$home_folder
ExperimentName <- params$experiment_name
date <- params$date
section_name <- params$section_name
network_of_interest <- params$network_of_interest
```


```{r prep, warning=FALSE, message=FALSE, echo = F}
# to render pngs in html
library(Cairo)
knitr::opts_chunk$set(fig.path = paste0(plotsdir, "/", tolower(gsub( x = params$complex_of_interest, pattern = " ", replacement = '_')), "/"), dev=c("CairoPNG", "pdf"))
print(params)
```
 
# Genes Clustered Together in UMAP

## UMAP

```{r highlighted_umap, warning = F, echo = F, fig.width = 6, fig.height = 6 }

ggplot(umap_coords, aes(x = umap_1, y = umap_2, 
                        col = ifelse(umap_1 < 1.75 & umap_1 > 1 & umap_2 > 2 & umap_2 < 2.75, "y", "n"))) + 
  xlab("UMAP_1") + ylab("UMAP_2") + 
  scale_color_manual(values = c('y' = 'red', 'n' = 'gray')) + 
  geom_point(size = 0.4) + theme_bw()  + theme(legend.position = 'none')

```
 

# Zoom In

```{r zoomed_in_umap, warning = F, echo = F, fig.width = 6, fig.height = 6 }

ggplot(umap_coords, aes(x = umap_1, y = umap_2, 
                        label = gene)) + 
  xlab("UMAP_1") + ylab("UMAP_2") + 
  xlim(c(1, 1.75)) + ylim(c(2, 2.75)) + 
  ggrepel::geom_text_repel(max.overlaps = Inf) + 
  geom_point() + theme_bw() 

```

All genes in the UMAP cluster:

```{r, echo = F}

genes_of_interest <- c("RPS27L", "TRAIP1", "BBC3", "BBS9", "BAX", "DDB2", "FDXR", "MDM2", "TIGAR", "TNFRSF10B", "IKBIP", "RRM2B", "DUSP14", "TM7SF3", "UFM1")
print("Downstream genes in UMAP cluster:")
print(genes_of_interest)

```

# Add in Highly Correlated Genes

## Distribution of Correlations

```{r umap_cluster_heatmap, echo = F, fig.width = 6, fig.height=6}

df4heatmap <- magpie_lfcs_by_downstream[genes_of_interest] %>%
  bind_rows() %>% as.data.frame() %>%
  dplyr::select(c("target", "downstream_gene_name", "lfc")) %>%
  reshape2::dcast(target ~ downstream_gene_name, value.var = 'lfc') %>%
  column_to_rownames('target') %>%
  cor()
print(my_heatmap(df4heatmap,
           treeheight_row = 0, treeheight_col = 0, min_c = -1, max_c = 1))

```

Distribution of Correlations:

```{r clust_cor_dist, echo = F, fig.width = 7, fig.height = 4}
correlated_genes <- filter(downstream_downstream_cor,
                           (downstream_gene_1 %in% genes_of_interest & downstream_gene_2 %in% genes_of_interest) & downstream_gene_1 < downstream_gene_2 )

dens <- density(correlated_genes$cor)
mean_cor = optimize(approxfun(dens$x, dens$y), interval = c(0.2, 0.3))$min
ggplot(correlated_genes, aes(x = cor)) + 
  xlab("Correlation within UMAP Cluster") + ylab("") + 
  geom_density() + 
  geom_vline(xintercept = mean_cor, lty = 3, col = 'red') + 
  annotate("text", x=mean_cor, y=0.25, label= paste0("Correlation cutoff: ", round(mean_cor, 4))) + 
  theme_bw()
```

Add in some other genes that weren't in the UMAP but are still highly correlated:

```{r, echo = F}

correlated_genes <- filter(downstream_downstream_cor,
                           (downstream_gene_1 %in% genes_of_interest | downstream_gene_2 %in% genes_of_interest) & abs(cor) > mean_cor )
correlated_genes <- unique(c(correlated_genes$downstream_gene_1, correlated_genes$downstream_gene_2))
correlated_genes <-correlated_genes[which(!(correlated_genes %in% genes_of_interest))]
print(paste0("Adding ", length(correlated_genes), " genes:"))
print(correlated_genes)

```

Heatmap of current genes:

```{r heatmap_with_correlated_genes, fig.width = 6, fig.height = 6, echo = F}

df4heatmap <- magpie_lfcs_by_downstream[c(genes_of_interest, correlated_genes)] %>%
  bind_rows() %>% as.data.frame() %>%
  dplyr::select(c("target", "downstream_gene_name", "lfc")) %>%
  reshape2::dcast(target ~ downstream_gene_name, value.var = 'lfc') %>%
  column_to_rownames('target') %>%
  cor()
print(my_heatmap(df4heatmap,
           treeheight_row = 0, treeheight_col = 0, min_c = -1, max_c = 1))
```

# Genes that Up/Downregulate these genes

```{r, echo = F}
targets_that_de <- magpie_lfcs_by_downstream[c(genes_of_interest, correlated_genes)] %>%
  bind_rows() %>% as.data.frame() %>%
  filter(pval_adj < sig_pval_thresh & target != downstream_gene_name) %>%
  .$target %>% table()
targets_that_de <- data.frame(
  target = names(targets_that_de),
  num_targets_of_interest = as.numeric(targets_that_de),
  num_targets_total = target_meta$n_downstream_excl_target[match(names(targets_that_de), target_meta$gene)],
  is_tf = names(targets_that_de) %in% tfs$source
) %>%
  mutate(frac_targets_of_interest = num_targets_of_interest/num_targets_total) %>%
  filter(num_targets_total > 3) %>% 
  arrange(desc(frac_targets_of_interest)) %>%
  slice_max(frac_targets_of_interest, n = 10) %>% 
  dplyr::select(c("target",
                  "num_targets_of_interest",
                  "num_targets_total"))
names(targets_that_de) <- c("Target", "Apoptosis Genes Perturbed by Target", "Number of Downstream DEG")
knitr::kable(targets_that_de)

```

Adding the following genes:

```{r, echo = F}
targets_that_de <- targets_that_de$Target
print(paste0("Adding ", length(targets_that_de), " genes:"))
print(targets_that_de)
```

Also, I feel like it's worth adding in TP53.

Check if there any other up-stream genes:

```{r}

targets_that_de_2 <- magpie_lfcs_by_downstream[targets_that_de] %>%
  bind_rows() %>% as.data.frame() %>%
  filter(pval_adj < sig_pval_thresh & target != downstream_gene_name) %>%
  .$target %>% table()

```

# Final selection



```{r target_downstream_heatmap, message = F, warning = F, fig.width = 8, fig.height = 4, echo = F}

df4heatmap <- magpie_lfcs_by_downstream[c(genes_of_interest, correlated_genes)] %>% bind_rows() %>% as.data.frame() %>%
  filter(target %in% c(targets_that_de, "TP53")) %>%
  dplyr::select(c("target", "downstream_gene_name", "lfc")) %>%
  reshape2::dcast(target ~ downstream_gene_name) %>%
  column_to_rownames('target')
my_heatmap(df4heatmap, midpoint = 0, treeheight_row = 0, treeheight_col = 0)
genes_to_consider <-data.frame(
  gene = c(genes_of_interest, correlated_genes, targets_that_de, "TP53"),
  reason = c(rep("in umap cluster", length(genes_of_interest)),
             rep("highly correlated deg", length(correlated_genes)),
             rep("target that perturbs", length(targets_that_de)),
             "tp53"
  )
)
genes_to_consider <- genes_to_consider[!(duplicated(genes_to_consider$gene)),]
fwrite(genes_to_consider, paste0(outdir, '/', params$date, "_apoptosis_genes.tsv"), sep  = '\t')
```

## UMAP

```{r highlighted_umap_all, warning = F, echo = F, fig.width = 6, fig.height = 6 }

ggplot(umap_coords, aes(x = umap_1, y = umap_2, 
                        col = ifelse(gene %in% genes_to_consider$gene, 'y', 'n'))) + 
  xlab("UMAP_1") + ylab("UMAP_2") + 
  scale_color_manual(values = c('y' = 'red', 'n' = 'gray')) + 
  geom_point(size = 0.4) + theme_bw()  + theme(legend.position = 'none')

```

# Session info

```{r session_info}
sessionInfo()
```