---
title: "The Hypoxia Genes"
author: "Claudia Feng"
date:  "`r format(Sys.Date(),'%e %B, %Y')`"
output: BiocStyle::html_document
params:
  date: "2022-08-15"
  home_folder : "/lustre/scratch123/hgi/teams/parts/cf14/crispr_scrnaseq_hipsci/"
  section_name : "12c_highly_correlated_downstream_genes"
  complex_of_interest : "hypoxia_pathway"
  experiment_name : "Magpie"
---

```{r, echo = F}
HomeFolder <- params$home_folder
ExperimentName <- params$experiment_name
date <- params$date
section_name <- params$section_name
network_of_interest <- params$network_of_interest
```


```{r prep, warning=FALSE, message=FALSE, echo = F}
# to render pngs in html
library(Cairo)
knitr::opts_chunk$set(fig.path = paste0(plotsdir, "/", tolower(gsub( x = params$complex_of_interest, pattern = " ", replacement = '_')), "/"), dev=c("CairoPNG", "pdf"))
print(params)
```
 
# Genes Clustered Together in UMAP

## UMAP

```{r highlighted_umap, warning = F, echo = F, fig.width = 6, fig.height = 6 }

ggplot(umap_coords, aes(x = umap_1, y = umap_2, 
                        col = ifelse(umap_1 <0 & umap_2 > 4 , "y", "n"))) + 
  xlab("UMAP_1") + ylab("UMAP_2") + 
  scale_color_manual(values = c('y' = 'red', 'n' = 'gray')) + 
  geom_point(size = 0.4) + theme_bw()  + theme(legend.position = 'none')

```
 

## Zoom In

All genes in the UMAP cluster:

```{r, echo = F}

genes_of_interest <- c("BNIP3", "IGFBP2", "ALDOA", "LDHA", "PGK1", "FAM162A", "PGAM1", "ENO1", "TPI1", "C4orf3", "PKM", "GPI")
print("Downstream genes in UMAP cluster:")
print(genes_of_interest)

```

## Functional Enrichment

# Add in Highly Correlated Genes

## Distribution of Correlations

```{r umap_cluster_heatmap, echo = F, fig.width = 7, fig.height=6}

df4heatmap <- magpie_lfcs_by_downstream[genes_of_interest] %>%
  bind_rows() %>% as.data.frame() %>%
  dplyr::select(c("target", "downstream_gene_name", "lfc")) %>%
  reshape2::dcast(target ~ downstream_gene_name, value.var = 'lfc') %>%
  column_to_rownames('target') %>%
  cor()
print(my_heatmap(df4heatmap,
           treeheight_row = 0, treeheight_col = 0, min_c = -1, max_c = 1,
      border_col = NA))

```

# Genes that Up/Downregulate these genes

```{r, echo = F}
targets_that_de <- magpie_lfcs_by_downstream[genes_of_interest] %>%
  bind_rows() %>% as.data.frame() %>%
  filter(pval_adj < sig_pval_thresh & target != downstream_gene_name) %>%
  .$target %>% table()
targets_that_de <- data.frame(
  target = names(targets_that_de),
  num_targets_of_interest = as.numeric(targets_that_de),
  num_targets_total = target_meta$n_downstream_excl_target[match(names(targets_that_de), target_meta$gene)],
  is_tf = names(targets_that_de) %in% tfs$source
) %>%
  mutate(frac_targets_of_interest = num_targets_of_interest/num_targets_total) %>%
  filter(num_targets_total > 3) %>% 
  arrange(desc(frac_targets_of_interest)) %>%
  slice_max(frac_targets_of_interest, n = 4) %>% 
  dplyr::select(c("target",
                  "num_targets_of_interest",
                  "num_targets_total")) %>%
  dplyr::mutate(
    sensitivity = num_targets_of_interest/num_targets_total,
    precision = num_targets_of_interest/length(c(genes_of_interest, correlated_genes))
    ) %>%
  dplyr::filter(
    !(target %in% genes2exclude)
  )
names(targets_that_de) <- c("Target", "Glycolysis Genes Perturbed by Target", "Number of Downstream DEG")
to_save <- list(
  genes_of_interest = c(genes_of_interest, correlated_genes),
  targets_that_de = targets_that_de
)
saveRDS(to_save, paste0(outdir, '/', date, '_targets-that-cause-glycolysis.RDS'))
knitr::kable(targets_that_de)

```

Adding the following genes (adding in VHL manually):

```{r, echo = F}
targets_that_de <- unique(c("VHL", targets_that_de$Target))
print(paste0("Adding ", length(targets_that_de), " genes:"))
print(targets_that_de)
targets_that_de <- c("VHL", "ARNT", "HIF1A")
```

## Genes that Up/Down-regulate the Up/Down-regulators

Check for more up-stream genes.

```{r, echo = F}

targets_that_de_2 <- magpie_lfcs_by_downstream[targets_that_de] %>% 
  bind_rows() %>% as.data.frame() %>%
  filter(pval_adj < sig_pval_thresh & target != downstream_gene_name) %>%
  .$target %>% table()
print(targets_that_de_2)

```

No genes up-stream of VHL, HIF1A, ARNT, NOM1 or RASL10A.

# Final selection



```{r target_downstream_heatmap, message = F, warning = F, fig.width = 8, fig.height = 4, echo = F}

df4heatmap <- magpie_lfcs_by_downstream[c(genes_of_interest, correlated_genes)] %>% bind_rows() %>% as.data.frame() %>%
  filter(target %in% c(targets_that_de)) %>%
  dplyr::select(c("target", "downstream_gene_name", "lfc")) %>%
  reshape2::dcast(target ~ downstream_gene_name) %>%
  column_to_rownames('target')
my_heatmap(df4heatmap, midpoint = 0, treeheight_row = 0, treeheight_col = 0)
genes_to_consider <-data.frame(
  gene = c(genes_of_interest, correlated_genes, targets_that_de),
  reason = c(rep("in umap cluster", length(genes_of_interest)),
             rep("highly correlated deg", length(correlated_genes)),
             rep("target that perturbs", length(targets_that_de))
  )
) %>% mutate(
  off_target = gene %in% (filter(off_targets, target_gene %in% targets_that_de) %>% .$off_target_gene)
)
genes_to_consider <- genes_to_consider[!(duplicated(genes_to_consider$gene)),]
fwrite(genes_to_consider, paste0(outdir, '/', params$date, "_", params$complex_of_interest, ".tsv"), sep  = '\t')
```

## UMAP

```{r highlighted_umap_all, warning = F, echo = F, fig.width = 6, fig.height = 6}

ggplot(umap_coords, aes(x = umap_1, y = umap_2, 
                        col = ifelse(gene %in% genes_to_consider$gene, 'y', 'n'))) + 
  xlab("UMAP_1") + ylab("UMAP_2") + 
  scale_color_manual(values = c('y' = 'red', 'n' = 'gray')) + 
  geom_point(size = 0.4) + theme_bw()  + theme(legend.position = 'none')

```

# Total Causal Effects

```{r, fig.width = 24, fig.height=8}
targets_of_interest <- sort(unique(c(genes_of_interest, "VHL", "ARNT", "HIF1A"), names(magpie_lfcs_by_target)))
sig_effects <- magpie_lfcs_by_target[targets_of_interest] %>%
  bind_rows() %>% as.data.frame() %>%
  dplyr::filter(pval_adj < sig_pval_thresh & downstream_gene_name %in% genes_of_interest & target != downstream_gene_name) %>%
  dplyr::select(c("target", "downstream_gene_name", "lfc"))

g <- graph_from_data_frame(sig_effects)

set.seed(0)
layout_g <- layout_as_tree(g)
layout_g[,1] <- layout_g[,1]*(6/diff(range(layout_g[,1])))
layout_g[,2] <- layout_g[,2] - 0.5

## center the upregulating genes
layout_g[which(!(names(V(g)) %in% c("VHL", "ARNT", "HIF1A"))),1] <- layout_g[which(!(names(V(g)) %in% c("VHL", "ARNT", "HIF1A"))),1] - mean(layout_g[which(!(names(V(g)) %in% c("VHL", "ARNT", "HIF1A"))),1])

current_coords <- layout_g %>% as.data.frame() %>% mutate(gene = names(V(g)))
V(g)$fill_col = ifelse(names(V(g)) %in% c("VHL", "ARNT", "HIF1A"), target_col, downstream_col)
E(g)$line_col = ifelse(sig_effects$lfc < 0, down_regulated_blue, up_regulated_red)
plot(g,
     layout = layout_g,
     edge.size = 0.3, vertex.size = 30,
     vertex.label.family="Arial", vertex.label.font=2,
     vertex.color = V(g)$fill_col,
     edge.color = E(g)$line_col,
     vertex.label.color = 'white',
     rescale=FALSE)

to_save <- list(
  g = g,
  layout_g = layout_g
)
saveRDS(to_save, paste0(outdir, '/', params$complex_of_interest, '/', date, '_data4network.RDS'))
```


# Session info

```{r session_info}
sessionInfo()
```