---
title: "What predicts the number of downstream genes of a target?"
author: "Britta Velten"
date:  "`r format(Sys.Date(),'%e %B, %Y')`"
output: BiocStyle::html_document
params:
  utils_path: "../../io/Magpie_Utils.R"
  io_path: "../../io/Magpie_io.R"
  date: "2022-08-15"
  section_name: "10a_number_of_DEGs"
---

# Preparations
```{r prep, warning=FALSE, message=FALSE, echo = F}
# to render pngs in html
library(Cairo)
knitr::opts_chunk$set(fig.path = plotsdir, dev=c("CairoPNG", "pdf"))
```

# Pluripotency Markers

Pluripotency markers are obtained from: https://link.springer.com/article/10.1007/s12015-015-9622-8.

```{r, echo = F}
pluripotency_tfs <- list(
  'Core transcriptional circuitry' = c("POU5F1", 'SOX2', 
                      'MYC', 'MYCL',
                      'LIN28A', 'NANOG', 'SALL4', 'UTF1', 
                      'NR5A2'),
  'Apoptosis & cell cycle' = c('P53', 'CDKN1A', 'CDKN2', 'MDM2', 'CCND1'),
  'TGFb pathway' = c('TGFBR1', 'SNAI1', 'SNAI2'),
  'Chromatin remodeling' = c('RCOR2', "SIRT6", 'DOT1L', "MDB3"),
  'E-cadherin/b-catenin' = c("GJA1")
)
pluripotency_tf_labels <- data.frame(
  gene = unlist(pluripotency_tfs),
  label = rep(names(pluripotency_tfs), unlist(lapply(pluripotency_tfs, length)))
)
knitr::kable(pluripotency_tf_labels)
```

Total number of pluripotency factors:

```{r, echo = F}
dim(pluripotency_tf_labels)[1]
```

# Trans Effects

## Number of Trans Effects

```{r n_trans_effects, fig.width = 3, fig.height =4, echo = F, warning = F}

df4plot <- target_meta %>%
  dplyr::filter(gene %in% unlist(pluripotency_tfs))
ggplot(df4plot %>%
         mutate(n_downstream_excl_target = ifelse(is.na(n_downstream_excl_target), 0, n_downstream_excl_target)), aes(x = n_downstream_excl_target,y = reorder(gene, n_downstream_excl_target))) + 
  xlab("# of Trans Effects") + ylab("Target") + 
  geom_bar(stat = 'identity') + theme_bw()

df4plot %>% slice_max(n_downstream_excl_target, n =3) %>% dplyr::select(c("gene", "n_downstream_excl_target"))
```

Compare to power:

```{r n_trans_effects_vs_power, fig.width = 4.3, fig.height =4, echo = F, warning = F}
ggplot(df4plot, aes(x = n_cells, y = n_downstream_excl_target, label = gene)) + 
  ggrepel::geom_text_repel(max.overlaps = Inf, col = 'red') + 
  xlab("# of Perturbed Cells") + ylab("# of Trans Effects") + 
  geom_point() + theme_bw()
```

Compare to on-target expression:

```{r n_trans_effects_vs_on_target, fig.width = 8, fig.height =4, echo = F, warning = F}
p_on_target_lfc <- ggplot(df4plot, aes(x = target_lfc, y = n_downstream_excl_target, label = gene)) + 
  ggrepel::geom_text_repel(max.overlaps = Inf, col = 'red') + 
  xlab("LFC") + ylab("# of Trans Effects") + 
  geom_point() + theme_bw()
p_on_target_pval <- ggplot(df4plot, aes(x = -log10(pval_adj), y = n_downstream_excl_target, label = gene)) + 
  ggrepel::geom_text_repel(max.overlaps = Inf, col = 'red') + 
  xlab("-log10(pval_adj)") + ylab("# of Trans Effects") + 
  scale_x_log10() + 
  geom_point() + theme_bw()
ggarrange(p_on_target_lfc, p_on_target_pval)
```
## Number of Regulators

```{r n_regulators, fig.width = 4.3, fig.height =4, echo = F, warning = F}

ggplot(df4plot, aes(x = control_norm_expr, y = n_targets_that_de, label = gene)) + 
  ggrepel::geom_text_repel(max.overlaps = Inf, col = 'red') + 
  xlab("Control Expression") + ylab("# of Regulating genes") + 
  geom_point() + theme_bw()


```

# Correlation of Perturbation Profiles

## Number of Correlated Genes

```{r n_correlated_knockdowns, fig.width = 4.3, fig.height =4, echo = F, warning = F}

n_correlates <- target_target_cor_anno %>%
  dplyr::filter(gene_1 %in% unlist(pluripotency_tfs) & abs(cor_all) > sig_cor_thresh_target & gene_1 != gene_2)
n_correlates <- as.data.frame(table(n_correlates$gene_1))
knitr::kable(n_correlates)
ggplot(n_correlates, aes(x = Freq, y = reorder(Var1, Freq))) +
  xlab("# of Similar Knockdowns") + ylab('') +
  geom_bar(stat = 'identity') + theme_bw()

```

## Heatmap

First do just off plain Pearson's correlation

```{r, fig.height=5, fig.width = 6, echo = F, warning = F}
df4cor <- magpie_lfcs %>%
  dplyr::filter(target %in% unlist(pluripotency_tfs)) %>%
  dplyr::select(c("target", "downstream_gene_name", "lfc")) %>%
  reshape2::dcast(target ~ downstream_gene_name, value.var = 'lfc') %>%
  column_to_rownames('target')
df4heatmap <- cor(t(df4cor))
my_heatmap(df4heatmap, 
           min_c = -0.5, max_c = 0.5,
           border_col = NA,
           treeheight_row = 0, treeheight_col = 0)
```

Try weighted correlation?


```{r, fig.height=5, fig.width = 6, echo = F, warning = F}
df4cor_vals <- magpie_lfcs %>%
  dplyr::filter(target %in% unlist(pluripotency_tfs)) %>%
  dplyr::select(c("target", "downstream_gene_name", "lfc")) %>%
  reshape2::dcast(target ~ downstream_gene_name, value.var = 'lfc') %>%
  column_to_rownames('target')
df4cor_weights <- magpie_lfcs %>%
  dplyr::filter(target %in% unlist(pluripotency_tfs)) %>%
  dplyr::select(c("target", "downstream_gene_name", "v")) %>%
  reshape2::dcast(target ~ downstream_gene_name, value.var = 'lfc') %>%
  column_to_rownames('target')
df4heatmap <- cor(t(df4cor))
my_heatmap(df4heatmap, 
           min_c = -0.5, max_c = 0.5,
           border_col = NA,
           treeheight_row = 0, treeheight_col = 0)
```


## Number of Co-perturbed Genes

```{r n_co-perturbed_downstream fig.width = 4.3, fig.height =6, echo = F, warning = F}

perturbing_targets <- downstream_downstream_cor_anno %>%
  dplyr::filter(downstream_gene_1 %in% unlist(pluripotency_tfs) & abs(iPSC_cor) > 0.2 & downstream_gene_1 != downstream_gene_2)
n_coperturbed <- perturbing_targets %>% .$downstream_gene_1 %>%
  table() %>% as.data.frame()
ggplot(n_coperturbed, aes(x = Freq, y = reorder(`.`, Freq))) +
  xlab("# of Co-perturbed Genes") + ylab('') +
  geom_bar(stat = 'identity') + theme_bw()
knitr::kable(n_coperturbed )
```

# Get Genes to Cluster

## Heatmap of Pluripotency TFs

```{r heatmap_of_tfs_only, fig.width = 20, fig.height =4, echo = F, warning = F}
targets2consider <- unlist(pluripotency_tfs)
downstreams2consider <- magpie_lfcs %>%
  dplyr::filter(target %in% unlist(pluripotency_tfs) & pval_adj < 0.1 & 
                  abs(lfc) > 0.1 & target != downstream_gene_name) %>%
  .$downstream_gene_name %>% as.character() %>%
  table() %>% as.data.frame() %>%
  dplyr::filter(Freq >=1) %>%
  .$`.` %>% as.character()
downstreams2consider <- sort(unique(downstreams2consider, unlist(pluripotency_tfs)))

df4heatmap <- magpie_lfcs %>%
  dplyr::filter(target %in% targets2consider  & 
                  downstream_gene_name %in% downstreams2consider) %>%
  dplyr::select(c('target', 'downstream_gene_name', 'lfc', 'pval_adj')) %>%
  #mutate(lfc = ifelse(pval_adj < 0.1, lfc, 0)) %>%
  reshape2::dcast(target ~ downstream_gene_name, value.var = 'lfc') %>%
  column_to_rownames('target') %>%
  as.matrix()

row_anno <- target_meta %>%
  dplyr::filter(gene %in% row.names(df4heatmap)) %>%
  dplyr::select(c("gene", "pre_day10_lfc_mean")) %>%
  column_to_rownames('gene')
my_heatmap(df4heatmap,
           treeheight_row=0, treeheight_col = 0,
           min_c = -0.5, max_c = 0.5,
           border_col = NA, 
           annotation_row=row_anno)

```


## Heatmap of Pluripotency TFs + Correlated Knockdowns

```{r heatmap_of_tfs_and_correlates, fig.width = 20, fig.height =6, echo = F, warning = F}
new_targets <- target_target_cor_anno %>%
  dplyr::filter(gene_1 %in% unlist(pluripotency_tfs) & abs(cor_all) > 0.2 & gene_1 != gene_2)
targets2consider <- target_target_cor_anno %>%
  dplyr::filter(gene_1 %in% unlist(pluripotency_tfs) & abs(cor_all) > 0.2) %>%
  .$gene_2
downstreams2consider <- magpie_lfcs %>%
  dplyr::filter(target %in% unlist(pluripotency_tfs) & pval_adj < 0.1 & 
                  abs(lfc) > 0.1 & target != downstream_gene_name) %>%
  .$downstream_gene_name %>% as.character() %>%
  table() %>% as.data.frame() %>%
  dplyr::filter(Freq >=1) %>%
  .$`.` %>% as.character()
downstreams2consider <- sort(unique(downstreams2consider, unlist(pluripotency_tfs)))

df4heatmap <- magpie_lfcs %>%
  dplyr::filter(target %in% targets2consider  & 
                  downstream_gene_name %in% downstreams2consider) %>%
  dplyr::select(c('target', 'downstream_gene_name', 'lfc', 'pval_adj')) %>%
  #mutate(lfc = ifelse(pval_adj < 0.1, lfc, 0)) %>%
  reshape2::dcast(target ~ downstream_gene_name, value.var = 'lfc') %>%
  column_to_rownames('target') %>%
  as.matrix()

row_anno <- target_meta %>%
  dplyr::filter(gene %in% row.names(df4heatmap)) %>%
  dplyr::select(c("gene", "pre_day10_lfc_mean")) %>%
  column_to_rownames('gene')
my_heatmap(df4heatmap,
           treeheight_row=0, treeheight_col = 0,
           min_c = -0.5, max_c = 0.5,
           border_col = NA, 
           annotation_row=row_anno)

```


## Get Correlates

Remove anything without a lot of signal. Keep only pluripotency genes some signal (> 2 trans effects) or some number of highly correlated knockdowns ( > 1 other knockdown with R > 0.2).

```{r, fig.width = 22, fig.height =4.5, echo = F, warning = F}
pluripotency_gene_meta <- left_join(left_join(pluripotency_tf_labels, 
                                    n_correlates %>% dplyr::select(c('gene' = "Var1", 'n_correlated' ='Freq'))),
                                    target_meta %>% dplyr::select(c("gene", "n_downstream_excl_target", "n_cells"))) %>%
  mutate(n_correlated = ifelse(is.na(n_correlated), 0, n_correlated))
tfs_with_signal <- pluripotency_gene_meta %>%
  dplyr::filter(n_downstream_excl_target > 2 | n_correlated > 1)

correlated_knockdowns <- target_target_cor_anno %>%
  dplyr::filter(gene_1 %in% tfs_with_signal$gene & abs(cor_all) > sig_cor_thresh_target & gene_1 != gene_2 & !(gene_2 %in% targets2exclude)) %>% dplyr::select(c("gene_1", "gene_2", "cor_all"))

targets2consider <- correlated_knockdowns$gene_2
downstreams2consider <- magpie_lfcs %>%
  dplyr::filter(target %in% unlist(pluripotency_tfs) & pval_adj < 0.1 & 
                  abs(lfc) > 0.1 & target != downstream_gene_name) %>%
  .$downstream_gene_name %>% as.character() %>%
  table() %>% as.data.frame() %>%
  dplyr::filter(Freq >=1) %>%
  .$`.` %>% as.character()
downstreams2consider <- sort(unique(downstreams2consider, unlist(pluripotency_tfs)))

df4heatmap <- magpie_lfcs %>%
  dplyr::filter(target %in% targets2consider  & 
                  downstream_gene_name %in% downstreams2consider) %>%
  dplyr::select(c('target', 'downstream_gene_name', 'lfc', 'pval_adj')) %>%
  #mutate(lfc = ifelse(pval_adj < 0.1, lfc, 0)) %>%
  reshape2::dcast(target ~ downstream_gene_name, value.var = 'lfc') %>%
  column_to_rownames('target') %>%
  as.matrix()

row_anno <- target_meta %>%
  dplyr::filter(gene %in% row.names(df4heatmap)) %>%
  dplyr::select(c("gene", "pre_day10_lfc_mean")) %>%
  column_to_rownames('gene')
my_heatmap(df4heatmap,
           treeheight_row=0, treeheight_col = 0,
           min_c = -0.5, max_c = 0.5,
           border_col = NA, 
           annotation_row=row_anno)
```

# Write Table

```{r, echo = F}
pluripotency_tfs2aggregate <- list(
  "Chromatin remodeling" = c("RCOR2", "FAM122A", "KDM1A"),
  'Core transcriptional circuitry' = c("POU5F1", "NANOG", "SOX2"),
  "Cluster_1" = c("POU5F1", "SOX2", "NANOG", "SALL4"),
  "Cluster_2" = c("RCOR2", "UTF1", "GJA1", "CCND1", "SIRT6", "SNAI2", "MYCN", "SNAI1", "LIN28A", "MYC", "CDKN2A", "MYCL")
)
pluripotency_clusters_df <- data.frame(
  complex_name = names(pluripotency_tfs2aggregate),
  genes_in_complex = unlist(lapply(pluripotency_tfs2aggregate, FUN = function(x){paste(x, collapse = '_')}))
)
fwrite(pluripotency_clusters_df, paste0(outdir, date, '_tf_clusters.tsv'), sep = '\t')
```

# SessionInfo
```{r}
sessionInfo()
```



