---
title: "Variance due to CRISPR Efficacy"
author: "Britta Velten"
date:  "`r format(Sys.Date(),'%e %B, %Y')`"
output: BiocStyle::html_document
---

# Preparations
```{r prep, warning=FALSE, message=FALSE}
library(Cairo)
knitr::opts_chunk$set(fig.path = paste0(plotsdir, '/'), dev=c("CairoPNG", "pdf"))
```

# Genes Considered

Considered genes with high expression (>0.5 in control) and large downstream effect (> 100 trans effects). In total, we considered:

```{r, echo = F}
paste0(length(targets_of_interest$gene), " targets")
```


# UMAPs

## All Cells that we Considered

```{r umap_all_targets, echo = F, fig.width = 6, fig.height = 6, warning = F, message = F}
ggplot(umap_coords %>%
         dplyr::filter(target != NonTargetGeneName), aes(x = UMAP_1, y = UMAP_2, 
                        col = target)) + 
  geom_point( size =0.5) + 
  ylim(c(-4, 3)) + 
  geom_point(data =umap_coords %>%
               dplyr::filter(target == NonTargetGeneName), size = 0.2, alpha = 0.4, col = 'gray') +  
  theme_bw() + theme(legend.position = 'bottom') +
     guides(color = guide_legend(override.aes = list(size = 3), title = 'Target' ) )

```

## Just the Controls

Mainly for reference
```{r umap_of_control, echo=F, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}

ggplot(umap_coords %>%
         dplyr::filter(target ==NonTargetGeneName), aes(x = UMAP_1, y = UMAP_2)) + 
  geom_point( size =0.5, col = 'gray') + 
  ylim(c(-4, 3)) + theme_bw()

```

## UMAPs by target

```{r umaps_by_target, echo=F, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}

all_umaps_by_target <- mapply(tg = targets_of_interest$gene, FUN = function(tg){
  p <- ggplot(umap_coords %>%
         dplyr::filter(target ==tg), aes(x = UMAP_1, y = UMAP_2, 
                        col = on_target_expr)) + 
  geom_point( size =0.5) + 
  ylim(c(-4, 3)) + 
    scale_color_viridis_c(paste0("Expression of ", tg)) +
  geom_point(data =umap_coords %>%
               dplyr::filter(target == NonTargetGeneName), size = 0.2, alpha = 0.4, col = 'gray') +  
  theme_bw() + theme(legend.position = 'bottom') 
  return(p)
}, SIMPLIFY = F)

all_umaps_by_target

```


# Scatters

Say something more general

## Scatters per Knockdown

```{r on_target_vs_effect_size_scatters_by_target, echo=F, fig.height=9, fig.width=12, message=FALSE, warning=FALSE}

mean_effect_scatters_by_target <- mapply(tg = targets_of_interest$gene, FUN = function(tg){
  p <- ggplot(umap_coords %>%
         dplyr::filter(target == tg), aes(x = on_target_expr, y = abs_lfc_deg_mean)) + 
  xlab("On-target Expression") + ylab("Effect Size") + ggtitle(tg) + 
  geom_point(size = 0.4) + theme_bw()
  return(p)
}, SIMPLIFY = F)

ggarrange(plotlist = mean_effect_scatters_by_target, ncol = 4, nrow = 3)

```


## Correlation between On-Target and Mean Effect Size

```{r on_target_vs_effect_size_scatter, echo=F, fig.height=5, fig.width=5, message=FALSE, warning=FALSE}
ggplot(umap_coords %>%
         dplyr::filter(target != NonTargetGeneName), aes(x = on_target_expr, y = abs_lfc_deg_mean)) + 
  xlab("On-target Expression") + ylab("Effect Size") + 
  geom_point(size = 0.4) + #ggpointdensity::geom_pointdensity() + 
  geom_smooth(method = "lm") + 
  theme_bw()
```

# SessionInfo
```{r, echo = F}
sessionInfo()
```

