---
title: "Check the P-values"
author: "Britta Velten"
date:  "`r format(Sys.Date(),'%e %B, %Y')`"
output: BiocStyle::html_document
params:
  date: "2023-10-18"
---

The process was fairly complicated and there were a lot of bugs on first glance, so check that the results are in fact what you think they should be.

# Preparations
```{r prep, echo = F, warning=FALSE, message=FALSE}
# to render pngs in html
library(Cairo)
knitr::opts_chunk$set(fig.path = plotsdir, dev=c("CairoPNG", "pdf"))
```

# Check pairs

Did the pairs that we're meant to find p-values for actually get computed?

```{r, echo = F}
## check for missing permutations
most_recent <- unlist(lapply(strsplit(list.files(file.path(OutFolder, "15c_calc_var_expl_by_target", "15c_00_calc_var_expl_by_target_pick_pairs"), 
                                                 pattern = "_pairs_to_permute.tsv"), "_"), "[[", 1))
all_permutations2test <- fread(file.path(OutFolder, "15c_calc_var_expl_by_target", "15c_00_calc_var_expl_by_target_pick_pairs", paste0(most_recent, "_pairs_to_permute.tsv")))
permuted_files <- list.files(file.path(OutFolder, "15c_calc_var_expl_by_target", "15c_03_calc_var_expl_by_target_permute_lines_all_pairs/"), recursive = T)
permuted_pairs <- gsub(permuted_files, pattern = paste0(".*", most_recent, "_|[.]tsv[.]gz"), replacement = "")
permuted_pairs <- data.frame(
  target = unlist(lapply(strsplit(permuted_pairs, "_"), "[[", 1)),
  downstream_gene_name = unlist(lapply(strsplit(permuted_pairs, "_"), "[[", 2))
)
missing_pairs <- all_permutations2test %>%
  dplyr::filter(!(paste0(target, "_", downstream_gene_name) %in% paste0(permuted_pairs$target, "_", permuted_pairs$downstream_gene_name)))

if (dim(missing_pairs)[1] > 0){
  print("Missing some pairs!")
} else{
  print("All pairs have data!")
}
```


# Significant LFCs

Check the significant LFCs are reasonable.

## Distribution of Un-adjusted p-values

```{r lfc_pval_dist, fig.width = 8, fig.height=3, echo = F, warning = F, message = F}
ggplot(lfc_pvals, aes(x = pval)) + 
  xlab("Estimated p-value (unadjusted)") + ylab("# of Downstream x Target Pairs") + 
  geom_histogram() + theme_bw()
```
Some of these are driven by the targets that 
remove anything 

```{r lfc_pval_dist_with_min_line_thresh, fig.width = 8, fig.height=3, echo = F, warning = F, message = F}
ggplot(lfc_pvals %>%
         dplyr::filter(n_paired_lines >= min_n_paired_lines), aes(x = pval)) + 
  xlab("Estimated p-value (unadjusted)") + ylab("# of Downstream x Target Pairs") + 
  geom_histogram() + theme_bw()
```


## Number of Permutations Computed

```{r lfc_n_perms_computed, fig.width = 12, fig.height=8, echo = F, warning = F, message = F}
ggplot(lfc_pvals, aes(x = permutations_tested)) + 
  facet_wrap(~ n_paired_lines, scales = 'free_y') + 
  scale_y_log10() + 
  xlab("# of Permutations Computed") + ylab("# of Downstream x Target Pairs") + 
  geom_histogram() + theme_bw()
```


## # of Permutations Beating True Value

```{r lfc_n_perms_beating_true_value, fig.width = 12, fig.height=8, echo = F, warning = F, message = F}
ggplot(lfc_pvals, aes(x = perms_beating_true_value)) + 
  facet_wrap(~ n_paired_lines, scales = 'free_y') + 
  scale_y_log10() + 
  xlab("# of Permutations Beating True Value") + ylab("# of Downstream x Target Pairs") + 
  geom_histogram() + theme_bw()
```

# Significant Post-Knockdown Expression


## Distribution of Un-adjusted p-values

```{r post_kd_pval_dist_with_min_line_thresh, fig.width = 8, fig.height=3, echo = F, warning = F, message = F}
ggplot(post_kd_expr_pvals %>%
         dplyr::filter(n_paired_lines >= min_n_paired_lines), aes(x = pval)) + 
  xlab("Estimated p-value (unadjusted)") + ylab("# of Downstream x Target Pairs") + 
  geom_histogram() + theme_bw()
```

## Number of Permutations Computed

```{r post_kd_n_perms_computed, fig.width = 12, fig.height=8, echo = F, warning = F, message = F}
ggplot(post_kd_expr_pvals, aes(x = permutations_tested)) + 
  facet_wrap(~ n_paired_lines, scales = 'free_y') + 
  scale_y_log10() + 
  xlab("# of Permutations Computed") + ylab("# of Downstream x Target Pairs") + 
  geom_histogram() + theme_bw()
```

## # of Permutations Beating True Value

```{r post_kd_expr_n_perms_beating_true_value, fig.width = 12, fig.height=8, echo = F, warning = F, message = F}
ggplot(post_kd_expr_pvals, aes(x = perms_beating_true_value)) + 
  facet_wrap(~ n_paired_lines, scales = 'free_y') + 
  scale_y_log10() + 
  xlab("# of Permutations Beating True Value") + ylab("# of Downstream x Target Pairs") + 
  geom_histogram() + theme_bw()
```

# SessionInfo
```{r}
sessionInfo()
```



