---
title: "Performance of Guide Assignment"
author: "Claudia Feng"
output: BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{Vignette Title}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}  
---


```{r, echo=F, message=F, warning = F}
library(data.table)
library(Seurat)
library(dplyr)
library(reshape2)
library(ggplot2)
library(Cairo)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path = plotsdir, dev=c("CairoPNG", "pdf"))
```


# CRISPR Marker Sequences

## Correlation between Counts

```{r count_correlation, fig.width = 5.5, fig.height=5, echo = F}

df4correlation <- left_join(cell_metadata, crispr_counts_df,
                            by = c("V1" = 'id'))
df4correlation$num_umis[is.na(df4correlation$num_umis)] <- 0
crispr_counts <- apply(select(df4correlation, all_of(crispr_sequences)), 1, sum)
df4correlation <- df4correlation[crispr_counts > 0,]
## correlation between sequence
sequence_coexpression <- select(df4correlation, c("nCount_guides", all_of(crispr_sequences)))
colnames(sequence_coexpression) <- gsub(gsub(colnames(sequence_coexpression), pattern = "_counts", replacement = ""),
     pattern = "nCount_guides", replacement = "Guide UMIs")
sequence_coexpression <- cor(sequence_coexpression)
my_heatmap(sequence_coexpression, min_c = -1, max_c = 1, treeheight_row = 0, treeheight_col = 0)
```


## Correlation Between Activity Levels and CellRanger Counts

Only done for Pica.

```{r crispr_counts_vs_activity, fig.width = 10, fig.heigth = 5, echo = F}
if (ExperimentName == "Pica"){
  df4plot <- line_metadata
df4plot <- left_join(df4plot, crispr_sequence_stats, by = c("cell_line" = "line"))
df4plot <- select(df4plot, c("cell_line", "dCas9", "mScarlet", "sequence", "frac_nonzeros")) %>%
  filter(sequence %in% c("dCas9-KRAB-MeCP2", "mScarlet"))

p1 <- ggplot(df4plot %>% filter(sequence == "mScarlet"), aes(x = mScarlet, y = frac_nonzeros, label = cell_line)) + 
  ggrepel::geom_text_repel() + 
  ylab("Fraction of Cells with Non-zero Counts") + 
  geom_point() + theme_bw()
p2 <- ggplot(df4plot %>% filter(sequence == "dCas9-KRAB-MeCP2"), aes(x = dCas9, y = frac_nonzeros, label = cell_line)) + 
  ggrepel::geom_text_repel() + 
  ylab("") + 
  geom_point() + theme_bw()
ggarrange(p1, p2)

}
```


## Counts vs. Assignment Status

Only done for Pica.

```{r counts_vs_status, fig.width = 15, fig.height=3, echo = F}
if (ExperimentName == "Pica"){
df4plot <- line_metadata
names(df4plot)[1] <- "line"
df4plot <- left_join(df4plot, crispr_sequence_stats)

## fraction of zero counts for assigned vs. unassigned

## do assigned cells have more dCas9/BSD/etc.
df4plot <- select(df4plot, c("frac_nonzeros_assigned", "frac_nonzeros_unassigned", "line", "sequence"))
names(df4plot) <- c("Assigned", "Unassigned", "Cell Line", "CRISPR Marker Sequence")
df4plot <- melt(df4plot, id = c("Cell Line", "CRISPR Marker Sequence"))
names(df4plot)[3] <- "Assignment Status"
ggplot(df4plot, aes(x = value, y = ..scaled..,col = `Assignment Status`)) + 
  facet_wrap(~`CRISPR Marker Sequence`, nrow = 1) + 
  xlab("Fraction of Cells with No Counts") + ylab("") + 
  geom_density(alpha = 0.5) + theme_bw()
}
```



# Target Gene Downregulation

## The Volcano Plot

```{r volcano_plot, fig.width=6, fig.height=4, echo = F, warning = F}
ggplot(target_df, aes(x=target_lfc, y= -log10(pval_adj))) + 
  geom_point(size = 0.4) + theme_classic() +
  geom_vline(xintercept = 0, lty = "dashed", col = "grey") + 
  xlab("Target expression log-foldchange") + ylab("Significance") + #(-log10(padj))
  guides(col="none")

```

## Scatter Plot with Background

```{r target_lfc_scatter, fig.width=4, fig.height=4, echo = F, warning = F}

df4plot <- lfcs %>%
  dplyr::select(c("target", "downstream_gene_name", "control_norm_expr", "lfc")) %>%
  mutate(on_target = ifelse(target == downstream_gene_name, "On-target effect", "Downstream Change"),
         post_kd_expr = control_norm_expr + lfc)
df4plot <- df4plot[order(df4plot$on_target),]
ggplot(df4plot, aes(x = control_norm_expr, y = post_kd_expr, col = on_target)) + 
  xlab("Mean Expression in Control Cells") + ylab("Post-Knockdown Expression") + 
  scale_color_manual("", values = c("On-target effect" = "red", "Downstream Change" = "gray") ) + 
  scale_alpha_manual("", values = c("On-target effect" = 1, "Downstream Change" = 0.4) ) + 
  geom_point( size = 0.4) + 
  geom_abline(slope = 1, lty = 3) + 
  theme_bw() + theme(legend.position = 'none')

```

## Confidence of Assignment vs. Target Gene Expression

```{r, echo = F, fig.width = 12, fig.height = 5, eval = T, echo = F, warning= F}

n_bins <- 10
cells_with_target_expr <- filter(on_target_expr, true_assignment_expr != -Inf)
guide_umi_bins <- floor((rank(cells_with_target_expr$num_umis) - 1)*n_bins/dim(cells_with_target_expr)[1]) + 1
#table(guide_umi_bins)
guide_umi_cutoffs <- sort(cells_with_target_expr$num_umis)[floor(1:n_bins*length(cells_with_target_expr$V1)/n_bins)]

## dCas9 vs. no dCas9
df4plot <-cells_with_target_expr %>%
  mutate(guide_umi_bin = guide_umi_bins) %>%
  filter(!(is.na(true_assignment_control_mean_expr))) %>%
  mutate(is_downregulated = ifelse(true_assignment_expr < true_assignment_control_mean_expr, "downregulated", 'not_downregulated')) %>%
  dplyr::select(c('dCas9_expression', 'guide_umi_bin', 'is_downregulated')) %>%
  table() %>% as.data.frame() %>%
  reshape2::dcast(dCas9_expression + guide_umi_bin ~ is_downregulated, value.var = 'Freq') %>%
  mutate(Freq = downregulated/(downregulated + not_downregulated))


## fraction of cells down-regulated with random assignment
frac_cells_downreg_rnd_assignment <- on_target_expr %>% filter(!is.infinite(random_assignment_expr) & !(is.na(random_assignment_control_mean_expr)))
frac_downregulated_random_assignment <- length(which(frac_cells_downreg_rnd_assignment$random_assignment_expr < frac_cells_downreg_rnd_assignment$random_assignment_control_mean_expr))/dim(frac_cells_downreg_rnd_assignment)[1]
                                                                                                    


levels(df4plot$guide_umi_bin) <- paste0(0:(n_bins - 1)*100/n_bins, "-", 1:n_bins*100/n_bins, "%", '\n> ', guide_umi_cutoffs, ' UMIs')

ggplot(df4plot, aes(x = guide_umi_bin, y = Freq, group=dCas9_expression, col = dCas9_expression)) + 
  ylab("Fraction of Cells with Down-regulated Target") + xlab("# Guide UMIs (Percentile)") + 
  ylim(c(min(c(df4plot$Freq, frac_downregulated_random_assignment)), 1)) + 
  scale_color_manual("Expression of dCas9-KRAB-MeCP2", values = c("Expressed" = "darkgreen", "Not Expressed" = 'red')) + 
  geom_line() + 
  geom_point() +
  geom_hline(yintercept = c(frac_downregulated_random_assignment, 1), lty = 3, col = 'black') + 
  theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

```



## Density Plots by Line

Done only for Pica.

```{r target_gene_downregulation_density, fig.width = 12, fig.height = 15, eval = F, echo = F}
cell_lines <- unique(target_df$cell_line)
plotlist <- mapply(l = cell_lines, FUN = function(l){
  df4plot <- target_df %>%
    mutate(to_color = ifelse(cell_line == l, l, ""))
  p <- ggplot(df4plot, aes(x=lfc, fill = as.character(cell_line == l))) + 
    scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "gray")) + 
    geom_density(alpha = 0.4) + theme_classic() +
    #facet_wrap(~ cell_line, ncol = 4) +
    geom_vline(xintercept = 0, lty = "dashed", col = "grey") + 
    xlab("Target expression log-foldchange") + ylab("Density")  + ggtitle(l) + #(-log10(padj))
    guides(col="none") + theme(legend.position = 'none')
  return(p)
}, SIMPLIFY = F)
ggarrange(plotlist = plotlist, ncol = 4, nrow = 5)
```

## Ranking the lines by target gene down-regulation

Done only for Pica. 

```{r line_ranking, fig.width = 4, fig.height = 6, eval = F, echo = F}
split_target_df <- split.data.frame(target_df, f = target_df$target)
split_target_df <- mapply(split_target_df, FUN = function(df){
  #print(df$target[1])
  r <- rank(df$lfc)
  names(r) <- df$cell_line
  rtn <- data.frame(target = df$target[1],
                    cell_line = cell_lines) %>%
    mutate(on_target_ranking  = ifelse(cell_line %in% names(r), r[cell_line], 25))
                  
  return(rtn)
}, SIMPLIFY = F) %>% bind_rows() %>% as.data.frame()
split_target_df <- split.data.frame(split_target_df, f = split_target_df$cell_line)
mean_on_target_ranking <- lapply(split_target_df, FUN = function(df){mean(df$on_target_ranking)}) %>% unlist()
mean_on_target_ranking <- data.frame(
  cell_line = names(mean_on_target_ranking),
  mean_ranking = mean_on_target_ranking
)
ggplot(mean_on_target_ranking, aes(x = mean_ranking, y = reorder(cell_line, mean_ranking))) + 
  xlab('Rank of On-Target LFC') + ylab('') + 
  geom_bar(stat = 'identity') + theme_bw()


```





Done only for Pica. Need to use target-gene LFC, as number of DEG is dependent on the number of cells.

```{r predicting_line_efficacy, fig.width = 5, fig.height = 15, eval = F, echo = F}
df4plot <- left_join(line_metadata , mean_on_target_ranking)

p1 <- ggplot(df4plot, aes(x = dCas9, y = mean_ranking, label = cell_line)) + 
  ggrepel::geom_text_repel() + 
  xlab("dCas9 Activity") + ylab("Line Efficacy Ranking") + 
  geom_point() + theme_bw()
p2 <- ggplot(df4plot, aes(x = mScarlet, y = mean_ranking, label = cell_line)) + 
  ggrepel::geom_text_repel() + 
  xlab("mScarlet Activity") + ylab("Line Efficacy Ranking") + 
  geom_point() + theme_bw()
p3 <- ggplot(df4plot, aes(x = frac_cells_assigned, y = mean_ranking, label = cell_line)) + 
  ggrepel::geom_text_repel() + 
  xlab("Fraction of Cells Assigned") + ylab("Line Efficacy Ranking") + 
  geom_point() + theme_bw()
ggarrange(p1, p2, p3, ncol =1 , nrow = 3)

```



# Fraction of Assigned Cells

## Fraction of Assigned Cells

```{r fraction_of_assigned_cells_bar, fig.width = 3, fig.height = 5, eval = T, echo = F}
df4plot <- on_target_expr %>%
  mutate(Status = ifelse(target_gene_name == "unassigned", "Unassigned",
                         ifelse(target_gene_name == NonTargetGeneName, "Non-targeting Control",
                                "Perturbed")))

ggplot(df4plot, aes(x = Status)) + 
  ylab("# of Cells") + xlab('') + 
  scale_y_log10() + 
  geom_bar(stat = 'count') + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```



## Predicting Fraction of Assigned Cells based on UMIs

```{r umis_predicting_cell_status, fig.width = 5, fig.height=5, echo = F, warning = F}
df4glm <- left_join(cell_metadata, crispr_counts_df, by = c("V1" = "id")) 
df4glm <- df4glm %>%
  mutate(status = ifelse(target_gene_name == "unassigned", 0, 1)) %>%
  dplyr::select(c("status", "nCount_guides", crispr_sequences))
## normalize eveyrthing
df4glm <- data.frame(
  status = df4glm$status,
  apply(df4glm[,-1], 2, FUN = function(x){(x - mean(x))/sd(x)})
)
fit <- glm(status ~ nCount_guides + BFP + BSD + dCas9.KRAB.MeCP2 + mScarlet + PURO, family=binomial(link='logit'),
                   data = df4glm)
df4plot <- as.data.frame(summary(fit)$coefficients)[, c(1,4)]
names(df4plot) <- c("Coefficient", "pval")
df4plot <- df4plot[-1,] %>% rownames_to_column("xlab_annotation")
df4plot$padj <- p.adjust(df4plot$pval, method = "BH")
df4plot$xlab_annotation <- gsub(df4plot$xlab_annotation, pattern = "dCas9.KRAB.MeCP2", replacement = "dCas9-KRAB-MeCP2")

ggplot(df4plot, aes(x = xlab_annotation, y = Coefficient, fill = padj < 0.05)) + 
    ylab('GLM Coefficient') + xlab('') + ggtitle('') + 
    geom_text(aes(label=ifelse(padj<0.001,"***", ifelse(padj<0.01,"**", ifelse(padj<0.05,"*","")))), 
              vjust=0.6 - sign(df4plot$Coefficient)/2) + 
    scale_fill_manual(values = c("TRUE" = 'darkblue', "FALSE" = 'gray')) + 
    geom_bar(stat="identity") + theme_bw()+ 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), plot.margin = margin(1,1,0, 2, "cm"), legend.position = 'none')
 
```

## Fraction of Assigned Cells per Line

Only done for Pica.

```{r fraction_of_assigned_cells_per_line_bar, fig.width = 10, fig.height = 5, eval = T, echo = F}
if (ExperimentName == "Pica"){
  ggplot(line_metadata, aes(x = cell_line, y = frac_cells_assigned, fill = donor)) + 
  xlab('') + ylab("Fraction of Cells Assigned") + 
  geom_bar(stat = 'identity') + theme_bw() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
}
```

## Predicting the Fraction of Cells Assigned by Line

Only done for Pica.

```{r predicting_frac_cell_assigned, fig.width = 10, fig.height=8, eval = F, echo = F}
#how do we predict the crappy lines
p1 <- ggplot(line_metadata, aes(x = dCas9, y = frac_cells_assigned,
                          label = cell_line)) + 
  ggrepel::geom_label_repel() + 
  xlab("dCas9 Activity") + ylab("Fraction of Cells Assigned") + 
  geom_point() + theme_bw()
p2 <- ggplot(line_metadata, aes(x = mScarlet, y = frac_cells_assigned,
                          label = cell_line)) + 
  ggrepel::geom_label_repel() + 
  xlab("mScarlet Expression") + ylab("Fraction of Cells Assigned") + 
  geom_point() + theme_bw()
p3 <- ggplot(line_metadata, aes(x = pluripotency_score, y = frac_cells_assigned,
                          label = cell_line)) + 
  ggrepel::geom_label_repel() + 
  xlab("Pluripotency Score") + ylab("Fraction of Cells Assigned") + 
  geom_point() + theme_bw()
p4 <- ggplot(line_metadata, aes(x = novelty_score, y = frac_cells_assigned,
                          label = cell_line)) + 
  ggrepel::geom_label_repel() + 
  xlab("Novelty Score") + ylab("Fraction of Cells Assigned") + 
  geom_point() + theme_bw()

#pdf(paste0(plotsdir, "/", date, "_frac_cells_assigned.pdf"), width = 10, height = 8)
ggarrange(p1, p2, p3, p4, nrow = 2, ncol = 2)
#dev.off()

```


# Down-regulation vs. Effect Size

Done only for Pica.

# Session Info

Session Info